---
title: "Codelist Generator"
---

```{r setup, include=FALSE}
library(webexercises)
install.packages("CohortConstructor")
install.packages("omock")
library(CohortConstructor)
library(omock)
```

## What is a cohort in OMOP?

A cohort is the fundamental unit of analysis in OMOP observational research. Formally, a cohort is a set of persons who satisfy one or more inclusion criteria for a duration of time. In OMOP, every cohort is stored as a table with exactly four columns:

| Column | Description |
|----|----|
| `cohort_definition_id` | Integer identifying which cohort definition this row belongs to |
| `subject_id` | The `person_id` of the patient |
| `cohort_start_date` | The date the patient enters the cohort (Day 0 / index date) |
| `cohort_end_date` | The date the patient exits the cohort |

This simple structure is powerful: one patient can appear multiple times (e.g. multiple treatment episodes), and one table can hold many different cohort definitions simultaneously.

::: callout-note
## Two ways to build cohorts

**ATLAS**: graphical interface, produces CIRCE-BE JSON → SQL. Best for sharing definitions across a network via JSON export.

**CohortConstructor (R)**: programmatic pipeline using `dplyr`-style syntax. Best for complex multi-step cohort logic, sensitivity analyses, and automated pipelines.

Both produce the same four-column cohort table. They are complementary, not competing.
:::

------------------------------------------------------------------------

## Setup

```{r setup-packages, message=FALSE, warning=FALSE}
library(CDMConnector)
library(CohortConstructor)
library(omopgenerics)
library(omock)
library(dplyr)

# Create synthetic CDM
cdm <- omock::mockCdmReference() |>
  omock::mockPerson(nPerson = 2000) |>
  omock::mockObservationPeriod() |>
  omock::mockConditionOccurrence() |>
  omock::mockDrugExposure() |>
  omock::mockDeath()
```

------------------------------------------------------------------------

## Building a cohort from concept sets

The most common starting point is `conceptCohort()`: define a cohort by the presence of a concept (or concept set) in a given domain table.

```{r concept-cohort, eval=FALSE}
# Define concept sets as named lists of concept IDs
glp1_concepts <- list(
  semaglutide  = c(2200644),   # RxNorm ingredient
  liraglutide  = c(1474828),
  dulaglutide  = c(1551099),
  exenatide    = c(1544838),
  tirzepatide  = c(37003518)
)

t2dm_concepts <- list(
  type2_diabetes = c(201826)   # SNOMED concept + descendants via ATLAS
)

ap_concepts <- list(
  acute_pancreatitis = c(197456)  # SNOMED
)

# Build drug exposure cohorts (one per GLP-1 agent)
cdm <- CohortConstructor::conceptCohort(
  cdm         = cdm,
  conceptSet  = glp1_concepts,
  name        = "glp1_cohorts",   # name of the cohort table in the CDM
  domain      = "drug"            # look in drug_exposure
)

# Build condition cohorts
cdm <- CohortConstructor::conceptCohort(
  cdm         = cdm,
  conceptSet  = t2dm_concepts,
  name        = "t2dm_cohort",
  domain      = "condition"
)

cdm <- CohortConstructor::conceptCohort(
  cdm         = cdm,
  conceptSet  = ap_concepts,
  name        = "ap_cohort",
  domain      = "condition"
)
```

> **Our study:** We create five separate exposure cohorts, one per GLP-1 RA ingredient. The `conceptCohort()` function creates one cohort entry per drug exposure episode per patient. The next steps will apply the new-user and other inclusion criteria.

------------------------------------------------------------------------

## Applying inclusion criteria with `requireIsFirstEntry()`

The **new-user design** requires that patients have no prior use of any GLP-1 RA before their index date. We implement this in two steps: first restricting to first entry per person, then requiring no prior exposure.

```{r first-entry, eval=FALSE}
# Keep only the first exposure episode per patient
cdm$glp1_cohorts <- cdm$glp1_cohorts |>
  CohortConstructor::requireIsFirstEntry()
```

::: callout-tip
## Why first entry only?

Without this restriction, a patient prescribed semaglutide three times would contribute three index dates. This creates immortal time bias and violates the new-user design assumption. Always apply `requireIsFirstEntry()` before any other temporal criteria.
:::

------------------------------------------------------------------------

## Requiring prior observation (washout window)

Patients must have sufficient prior observation in the database to apply the washout period and measure baseline covariates:

```{r prior-obs, eval=FALSE}
cdm$glp1_cohorts <- cdm$glp1_cohorts |>
  CohortConstructor::requireObservationBeforeDrug(
    days = 365    # at least 365 days prior observation
  )
```

------------------------------------------------------------------------

## Requiring no prior exposure (washout)

```{r washout, eval=FALSE}
# Restrict to patients with no prior GLP-1 RA in 365 days before index
# This is the core of the new-user design
cdm$glp1_cohorts <- cdm$glp1_cohorts |>
  CohortConstructor::requirePriorDrugWashout(
    days = 365
  )
```

------------------------------------------------------------------------

## Requiring presence in another cohort (population criterion)

Patients must have a T2DM diagnosis on or before the index date:

```{r require-cohort, eval=FALSE}
cdm$glp1_cohorts <- cdm$glp1_cohorts |>
  CohortConstructor::requireCohortIntersect(
    cohortTable      = "t2dm_cohort",
    cohortId         = 1,
    window           = c(-Inf, 0),   # any time before or on index date
    intersections    = c(1, Inf),    # at least 1 intersection required
    name             = "glp1_cohorts"
  )
```

------------------------------------------------------------------------

## Restricting to a study period

```{r date-restrict, eval=FALSE}
cdm$glp1_cohorts <- cdm$glp1_cohorts |>
  CohortConstructor::requireInDateRange(
    dateRange = as.Date(c("2023-01-01", "2024-12-31"))
  )
```

------------------------------------------------------------------------

## Age restriction

```{r age-restrict, eval=FALSE}
cdm$glp1_cohorts <- cdm$glp1_cohorts |>
  CohortConstructor::requireAge(
    ageRange = list(c(18, 150))   # adults only
  )
```

------------------------------------------------------------------------

## Putting the full pipeline together

```{r full-pipeline, eval=FALSE}
# Full new-user cohort pipeline for GLP-1 RA agents
cdm <- CohortConstructor::conceptCohort(
  cdm        = cdm,
  conceptSet = glp1_concepts,
  name       = "glp1_cohorts",
  domain     = "drug"
) 

cdm$glp1_cohorts <- cdm$glp1_cohorts |>
  requireIsFirstEntry()             |>
  requireObservationBeforeDrug(days = 365) |>
  requirePriorDrugWashout(days = 365)      |>
  requireCohortIntersect(
    cohortTable   = "t2dm_cohort",
    cohortId      = 1,
    window        = c(-Inf, 0),
    intersections = c(1, Inf),
    name          = "glp1_cohorts"
  ) |>
  requireAge(ageRange = list(c(18, 150))) |>
  requireInDateRange(
    dateRange = as.Date(c("2023-01-01", "2024-12-31"))
  )
```

------------------------------------------------------------------------

## Checking cohort counts and attrition

After building the cohort, always inspect the attrition table — it shows how many patients were excluded at each step, which is essential for transparency and for detecting logic errors.

```{r attrition, eval=FALSE}
library(CohortCharacteristics)

# Cohort counts
attrition_result <- summariseCohortAttrition(cdm$glp1_cohorts)

# Visualise as a table
tableCohortAttrition(attrition_result)

# Visualise as a flowchart (for one cohort)
attrition_result |>
  filter(group_level == "semaglutide") |>
  plotCohortAttrition()
```

> **Our study:** A well-specified attrition table for the semaglutide cohort might look like this:
>
> | Step                          | Remaining N |
> |-------------------------------|-------------|
> | Initial drug exposure records | 45,210      |
> | First entry per person        | 12,847      |
> | ≥ 365 days prior observation  | 11,203      |
> | No prior GLP-1 RA (washout)   | 9,841       |
> | T2DM diagnosis before index   | 8,654       |
> | Age ≥ 18                      | 8,640       |
> | Index date 2023–2024          | 6,112       |
>
> If the washout step removes more than 60% of patients, this is worth investigating — it may indicate that your concept set is too broad and is capturing repeat prescriptions of the same episode, or that the database has many prevalent users.

```{r q1, echo=FALSE, results='asis'}
cat(
  "**Quick check:** Your attrition table shows 10,000 patients after `requireIsFirstEntry()`",
  "but only 200 after `requirePriorDrugWashout(days = 365)`. What does this most likely indicate?\n\n",
  mcq(c(
    "The drug is rarely prescribed in this database",
    answer = "Most drug users in this database are prevalent users, not new users — they had prior prescriptions in the year before their first entry",
    "The washout period is too short",
    "There is a bug in the CohortConstructor code"
  )), "\n\n"
)
```

`r hide("Explanation")` When the vast majority of patients are excluded by the washout criterion, it means most patients who appear in the database with this drug already had prior exposure — i.e. they are **prevalent users**. This is common for drugs that have been on the market for many years (like liraglutide since 2009 or exenatide since 2006). The new-user design deliberately excludes these patients to avoid prevalent-user bias, but it does reduce your study population substantially. This is a design trade-off, not a bug. `r unhide()`

------------------------------------------------------------------------

## Exporting cohort definitions

```{r export, eval=FALSE}
# Export cohort table to a local file for sharing or archiving
cohort_data <- cdm$glp1_cohorts |> collect()
write.csv(cohort_data, "results/glp1_cohorts.csv", row.names = FALSE)

# Disconnect
CDMConnector::cdmDisconnect(cdm)
```

## References

::: {#refs}
:::
