---
title: "Cohort Construction"
---

```{r setup, include=FALSE}
library(webexercises)
install.packages("CohortConstructor")
install.packages("omock")
library(CohortConstructor)
library(omock)
```

## What is a cohort in OMOP?

A cohort will be the fundamental unit of analysis in our studies. It is a set of persons who satisfy one or more inclusion criteria for a duration of time. In OMOP, every cohort is stored as a table with exactly four columns:

| Column | Description |
|----|----|
| `cohort_definition_id` | Integer identifying which cohort definition this row belongs to |
| `subject_id` | The `person_id` of the patient |
| `cohort_start_date` | The date the patient enters the cohort (Day 0 / index date) |
| `cohort_end_date` | The date the patient exits the cohort |

In this structure, one patient can appear multiple times (e.g. multiple treatment episodes), and one table can hold many different cohort definitions simultaneously.

::: callout-note
## The two ways to build cohorts

[**ATLAS**](atlas.qmd): it has the graphical interface, and produces a JSON file that can be exported. Suitable for most studies.

**CohortConstructor (R)**: R package developed by OHDSI. Best for complex multi-step cohort logic, sensitivity analyses, and automated pipelines.
:::

## Setup

In this example, we better work with the Eunomia dataset as an example instead of the omock data, because the number and variety of drug exposures available in the synthetic data are more broad in the Eunomia dataset. You can also download example datasets from the `omock` package, like the `GiBleed` dataset. The documentation of the package is [here](https://ohdsi.github.io/omock/index.html).

```{r setup-packages, message=FALSE, warning=FALSE}
library(CodelistGenerator)
library(CDMConnector)
library(omopgenerics)
library(omock)
library(dplyr)
library(duckdb)

# With Eunomia
requireEunomia()
db <- DBI::dbConnect(duckdb::duckdb(), dbdir = eunomiaDir())
cdm <- cdmFromCon(db, 
                  cdmSchema = "main", 
                  writeSchema = "main", 
                  writePrefix = "greg_") # change the prefix at your convenience

semaglutide <- getDrugIngredientCodes(cdm = cdm,
                                      name = "semaglutide")

t2dm <- getCandidateCodes(
  cdm = cdm,
  keywords = c("type 2 diabetis mellitus", "T2DM")
)

acute_pancreatitis <- getCandidateCodes(
  cdm = cdm,
  keywords = c("Acute pancreatitis")
)
```

::: callout-note
If the CDM you use does not have the `concept id` that allow you to do the exercises, please, try another CDM or use another substance/disease!
:::

------------------------------------------------------------------------

## Example pipeline

In this tutorial we will provide an example on how to build the cohorts for our study with `CohortConstructor`, but further documentation can be found in the [package documentation](https://ohdsi.github.io/CohortConstructor/), with all available functions in its [reference index](https://ohdsi.github.io/CohortConstructor/reference/index.html) and in [this other tutorial](https://dpa-pde-oxford.quarto.pub/cohortconstructorinternaltrainingsep25/#/cohortconstructor) or [this one](https://dpa-pde-oxford.quarto.pub/cohortconstructorrmedicine/#/cohortconstructor).

### 1. Building a cohort from concept sets

The most common starting point is `conceptCohort()`: define a cohort by the presence of a concept (or concept set) in a given domain table. Remember that we have our concept sets already from the [Codelist Generator](codelistgenerator.qmd) section!

```{r concept-cohort, eval=FALSE}
library(CohortConstructor)
  
# Build drug exposure cohorts (one per GLP-1 agent)
cdm$semaglutide_cohort <- conceptCohort(
  cdm         = cdm,
  conceptSet  = semaglutide,
  name        = "semaglutide_cohort",   # name of the cohort table in the CDM
  exit = "event_end_date",
  overlap = "merge")


# Build condition cohorts
cdm$t2dm_cohort <- conceptCohort(
  cdm         = cdm,
  conceptSet  = t2dm,
  name        = "t2dm_cohort"
)

cdm$ap_cohort <- conceptCohort(
  cdm         = cdm,
  conceptSet  = acute_pancreatitis,
  name        = "ap_cohort"
)


```

> **Our study:** We create five separate exposure cohorts, one per GLP-1 RA ingredient. The `conceptCohort()` function creates one cohort entry per drug exposure episode per patient. The next steps will apply the new-user and other inclusion criteria.

### 2. Applying inclusion criteria with `require()`

The **new-user design** requires that patients have no prior use of any GLP-1 RA before their index date. We implement this in two steps: first restricting to first entry per person, then requiring no prior exposure.

```{r first-entry, eval=FALSE}
cdm$semaglutide_cohort <- cdm$semaglutide_cohort |>
  requireIsFirstEntry() |> # Keep only the first exposure episode per patient
  requireObservationBeforeDrug(
    days = 365    # at least 365 days prior observation
  ) |>
  requirePriorDrugWashout(
    days = 365 # Restrict to patients with no prior semaglutide in 365 days before index
  ) |>
  requireAge(
    ageRange = list(c(18, 150)) # Restrict to only adults 
  )
  requireInDateRange(
    dateRange = as.Date(c("2023-01-01", "2025-12-31")) # Restrict to the study period
  ) 
```

::: callout-important
You need to have the 365 days observation period prior to the index date if you want to be able to apply the washout period of 365 days for new use! If you do not have this previous period of observation you are not able to discern whether a previous use has happened in these days.
:::

### 3. Requiring presence in another cohort

Patients must have a T2DM diagnosis on or before the index date:

```{r require-cohort, eval=FALSE}
cdm$semaglutide_cohort <- cdm$semaglutide_cohort |>
  requireCohortIntersect(
    cohortTable      = "t2dm_cohort",
    cohortId         = 1,
    window           = c(-Inf, 0),   # any time before or on index date
    intersections    = c(1, Inf),    # at least 1 intersection required
    name             = "semaglutide_cohort"
  )
```

::: callout-tip
## The order matters!

The order in which you introduce the inclusion criteria will have an impact on your final cohort. When you plot the attrition of the cohort building process, you will see how your sample of patients becomes increasingly restricted by your inclusion criteria. You might decide to be more strict or loose in some criteria depending on that! Try to start by more general rules and become more precise in the end of the pipeline. Or try to implement your most important restriction criteria first.
:::

```{r cohortconstr-q, echo=FALSE, results='asis'}
cat(
  "Do you think that the order that we followed for the construction of the cohort of semaglutide is correct?",
  "\n\n",
  mcq(c(
    "Yes, from general to specific",
    answer = "Yes, from unavoidable to changeable",
    "No, you should do the most general criteria first, like date range",
    "Yes, but I do not understand the order"
  )),
  "\n\n"
)
```

`r hide("Explanation")` The order we applied here is from the most unavoidable to the least. We want to define a new-user cohort, therefore restricting to first entry and making sure that the first entry is well-captured is a must. If we would like incident entry and not first entry, we should just remove the `requireIsFirstEntry()` command but all other restrictions apply. Finally, we can more easily adapt the study period than change the age restriction for our study, because this could also be drug/condition dependant. `r unhide()`

### 4. Finishing details

Here we created simple cohorts that exit at the end of an exposure or observation. We might want to specify other reasons for exit, such as entering another cohort, switching exposures, or even death. You can also do that and be sure to create properly your cohort definitions before continuing with the analyses.

### 5. Checking cohort counts and attrition

After building the cohort, we can inspect the attrition table, as it shows how many patients were excluded at each step, which is essential for transparency and for detecting logic errors. You can do that running diagnostics on your cohorts, as you will see in the [Diagnostics](diagnostics.qmd) section.

::: callout-tip
Always check your attrition table or plot! And make sure to run the appropriate [diagnostics](diagnostics.qmd) to your cohorts!
:::

### 6. Exporting cohort definitions

```{r export, eval=FALSE}
# Export cohort table to a local file for sharing or archiving
cohort_data <- cdm$semaglutide_cohort |> collect()
write.csv(cohort_data, "results/semaglutide_cohort.csv", row.names = FALSE)

# Disconnect
dbDisconnect(con)
```
