---
title: "Diagnostics"
format: html
---

```{r setup, include=FALSE}
library(webexercises)
install.packages("DBI")
install.packages("CDMConnector")
install.packages("duckdb")
install.packages("PhenotypeR")
install.packages("remotes")
# remotes::install_github("OHDSI/CohortDiagnostics")
library(DBI)
library(CDMConnector)
library(duckdb)
library(PhenotypeR)
library(remotes)
# library(CohortDiagnostics)
```

## Why to run diagnostics before analysis?

Running an analysis on a poorly specified cohort is worse than running no analysis at all — it produces numbers that look credible but are wrong. Diagnostic tools exist to systematically surface problems with your phenotype algorithms and cohort logic **before** you commit to an analytical result.

In the OHDSI/DARWIN EU ecosystem, the diagnostic workflow centres on two tools: **CohortDiagnostics** for population-level phenotype evaluation, and **CohortCharacteristics** for detailed baseline description. Think of these as your study's quality control layer.

------------------------------------------------------------------------

## Setup
In this example, we better work with the Eunomia dataset as an example instead of the omock data, because the number and variety of drug exposures available in the synthetic data are more broad in the Eunomia dataset. You can also download example datasets from the omock package, like the GiBleed dataset. The documentation of the package is here.

``` {r setup-packages, message=FALSE, warning=FALSE}
library(CDMConnector)
library(omopgenerics)
library(omock)
library(dplyr)
library(duckdb)

# Create synthetic CDM
cdm <- omock::mockCdmReference() |>
  omock::mockPerson(nPerson = 2000) |>
  omock::mockObservationPeriod() |>
  omock::mockConditionOccurrence() |>
  omock::mockDrugExposure() |>
  omock::mockDeath()

# Download mock datasets from omock
downloadMockDataset(datasetName = "GiBleed", path = NULL, overwrite = NULL)

# With Eunomia
requireEunomia()
db <- DBI::dbConnect(duckdb::duckdb(), dbdir = eunomiaDir())
cdm <- cdmFromCon(db, 
                  cdmSchema = "main", 
                  writeSchema = "main", 
                  writePrefix = "greg_") # change the prefix at your convenience

# ------------------------------------------------------------------------------------
# Remember that we previously created semaglutide_cohort, t2dm_cohort and ap_cohort!
# Go to the Cohort Constructor section if you need more info.
```

------------------------------------------------------------------------

## Example pipeline

In this tutorial we will provide an example on how to build the concept sets for our study with CodelistGenerator, but further documentation can be found in the [package documentation](https://darwin-eu.github.io/CodelistGenerator/), with all available functions in its [reference index](https://darwin-eu.github.io/CodelistGenerator/reference/index.html) and in [this other tutorial](https://dpa-pde-oxford.quarto.pub/codelistgenerator/#/title-slide).

1\. Cohort counts and attrition

The first diagnostic is always: how many patients are in the cohort, and how many were lost at each step?

```{r attrition, eval=FALSE}
# Cohort entry counts
count_result <- summariseCohortCount(cdm$glp1_cohorts)
tableCohortCount(count_result)

# Attrition (if you built the cohort with CohortConstructor)
attrition_result <- summariseCohortAttrition(cdm$glp1_cohorts)
tableCohortAttrition(attrition_result)

# Plot the attrition flowchart for one cohort
attrition_result |>
  filter(group_level == "semaglutide") |>
  plotCohortAttrition()
```

>
> **Our study:** A well-specified attrition table for the semaglutide cohort might look like this:
>
> | Step                          | Remaining N |
> |-------------------------------|-------------|
> | Initial drug exposure records | 45,210      |
> | First entry per person        | 12,847      |
> | ≥ 365 days prior observation  | 11,203      |
> | No prior GLP-1 RA (washout)   | 9,841       |
> | T2DM diagnosis before index   | 8,654       |
> | Age ≥ 18                      | 8,640       |
> | Index date 2023–2024          | 6,112       |
>
> If the washout step removes more than 60% of patients, this is worth investigating — it may indicate that your concept set is too broad and is capturing repeat prescriptions of the same episode, or that the database has many prevalent users.
>
> ------------------------------------------------------------------------


```{r q1, echo=FALSE, results='asis'}
cat(
  "**Quick check:** Your attrition table shows 10,000 patients after `requireIsFirstEntry()`",
  "but only 200 after `requirePriorDrugWashout(days = 365)`. What does this most likely indicate?\n\n",
  mcq(c(
    "The drug is rarely prescribed in this database",
    answer = "Most drug users in this database are prevalent users, not new users — they had prior prescriptions in the year before their first entry",
    "The washout period is too short",
    "There is a bug in the CohortConstructor code"
  )), "\n\n"
)
```

`r hide("Explanation")` When the vast majority of patients are excluded by the washout criterion, it means most patients who appear in the database with this drug already had prior exposure — i.e. they are **prevalent users**. This is common for drugs that have been on the market for many years (like liraglutide since 2009 or exenatide since 2006). The new-user design deliberately excludes these patients to avoid prevalent-user bias, but it does reduce your study population substantially. This is a design trade-off, not a bug. `r unhide()`
## 2. Index event breakdown

The index event breakdown tells you which specific concepts are driving cohort entry. For a drug cohort, this shows which branded products or formulations account for the most index dates.

```{r index-breakdown, eval=FALSE}
# Summarise the concept codes that appear on the index date
# (requires CohortDiagnostics — shown here conceptually)

# In CohortCharacteristics, use large scale characteristics
# restricted to the index date window
index_result <- cdm$glp1_cohorts |>
  summariseLargeScaleCharacteristics(
    window            = list(c(0, 0)),       # on index date only
    eventInWindow     = "drug_exposure",
    minimumFrequency  = 0.005                # only show if >0.5% of cohort
  )

tableTopLargeScaleCharacteristics(index_result)
```

::: callout-note
## What to look for

For the semaglutide exposure cohort, you expect to see Ozempic (injectable) and Rybelsus (oral) as the dominant index concepts. If you see unexpected concepts — for example, a weight-loss branded product like Wegovy being excluded when you wanted to include it — your concept set needs revision.
:::

------------------------------------------------------------------------

## 3. Time distributions

How long do patients have prior observation before the index date? And how long do they remain in the cohort post-index?

```{r time-dist, eval=FALSE}
# Summarise cohort entry timing
timing_result <- summariseCohortTiming(cdm$glp1_cohorts)
tableCohortTiming(timing_result)

plotCohortTiming(
  timing_result,
  uniqueCombinations = TRUE,
  timeScale          = "years"
)
```

> **Our study:** The distribution of prior observation time should show a minimum of 365 days (our washout + observation requirement). If you see patients with fewer than 365 days, your `requireObservationBeforeDrug()` criterion was not applied correctly. The distribution of post-index time tells you how much follow-up you have — if most patients have less than 6 months of follow-up, your 3-year HR estimate will be very imprecise.

------------------------------------------------------------------------

## 4. Baseline characteristics

The core diagnostic for confounding: are the five GLP-1 RA cohorts comparable at baseline, or do they differ systematically in ways that could explain any observed difference in pancreatitis risk?

```{r characteristics, eval=FALSE}
# Summarise age, sex, and prior conditions
char_result <- summariseCharacteristics(
  cohort   = cdm$glp1_cohorts,
  ageGroup = list(c(18, 44), c(45, 64), c(65, 150)),
  tableIntersectCount = list(
    list(
      tableName  = "condition_occurrence",
      window     = c(-365, -1)
    )
  )
)

tableCharacteristics(char_result)

# Age distribution by cohort
char_result |>
  filter(variable_name == "Age") |>
  plotCharacteristics(
    plotType = "boxplot",
    colour   = "cohort_name"
  )
```

------------------------------------------------------------------------

## 5. Large-scale characteristics — the SMD plot

The standardised mean difference (SMD) is the standard metric for assessing covariate balance across cohorts. An SMD below 0.1 for a covariate is generally considered indicative of no meaningful imbalance.

```{r lsc, eval=FALSE}
# Large scale characteristics: all conditions, drugs, procedures
lsc_result <- cdm$glp1_cohorts |>
  summariseLargeScaleCharacteristics(
    window            = list(c(-365, -1)),    # 1 year prior to index
    eventInWindow     = c("condition_occurrence", "drug_exposure"),
    minimumFrequency  = 0.01                  # at least 1% prevalence
  )

# Comparison plot — SMD between cohorts
lsc_result |>
  plotComparedLargeScaleCharacteristics(
    reference = "semaglutide",
    colour    = "variable_level"
  )
```

> **Our study:** After running large-scale characteristics across all five GLP-1 RA cohorts, we expect some imbalances. Key ones to look for:
>
> - **Hypertriglyceridaemia** — if more prevalent in older-drug cohorts (exenatide, liraglutide), this is a confounder for the pancreatitis outcome
> - **Obesity** — tirzepatide may be preferentially prescribed to more obese patients, which itself increases pancreatitis risk
> - **Prior pancreatitis** — must be inspected carefully; any prior acute pancreatitis should be handled as an exclusion criterion or sensitivity analysis

```{r q1, echo=FALSE, results='asis'}
cat(
  "**Quick check:** You plot the SMD for 500 baseline covariates and find that",
  "hypertriglyceridaemia has an SMD of 0.32 between semaglutide and exenatide users.",
  "What does this tell you?\n\n",
  mcq(c(
    "The difference is small and can be ignored",
    answer = "There is meaningful imbalance — hypertriglyceridaemia should be included in the propensity score model as it is both a confounder and a risk factor for the outcome",
    "You should exclude all patients with hypertriglyceridaemia",
    "The cohort definitions are wrong and need to be redefined"
  )), "\n\n"
)
```

`r hide("Explanation")` An SMD of 0.32 is well above the conventional threshold of 0.1, meaning hypertriglyceridaemia is meaningfully more prevalent in one exposure group than the other. Since hypertriglyceridaemia is also an independent risk factor for acute pancreatitis, it meets the definition of a confounder — it is associated with both treatment assignment and the outcome. It must be included in the propensity score model or adjusted for in the outcome model. Excluding patients with hypertriglyceridaemia would change your target population and invalidate the generalisability of your results. `r unhide()`

------------------------------------------------------------------------

## 6. Cohort overlap

Do patients appear in more than one of our five exposure cohorts? This should be rare or impossible if your new-user + washout criteria are correctly specified — a patient should only initiate one GLP-1 RA first.

```{r overlap, eval=FALSE}
overlap_result <- summariseCohortOverlap(cdm$glp1_cohorts)
tableCohortOverlap(overlap_result)
plotCohortOverlap(overlap_result, uniqueCombinations = TRUE)
```

::: callout-tip
## If overlap is high

High overlap between exposure cohorts means your cohort exit criteria are not correctly excluding patients who switch between agents. For example, if a patient switches from liraglutide to semaglutide, they should exit the liraglutide cohort at the switch date and — if their second entry passes the washout criterion — enter the semaglutide cohort. Failure to handle this correctly leads to immortal time and misclassification bias.
:::

------------------------------------------------------------------------

## Exporting diagnostics results

```{r export, eval=FALSE}
# Combine all diagnostic results
all_diagnostics <- omopgenerics::bind(
  count_result,
  attrition_result,
  char_result,
  lsc_result,
  overlap_result
) |>
  omopgenerics::suppress(minCellCount = 5)  # privacy protection

# Export for sharing with study team
write.csv(all_diagnostics, "results/diagnostics.csv", row.names = FALSE)
```
