---
title: "Drug Utilisation"
format: html
---

```{r setup, include=FALSE}
library(webexercises)
install.packages("DrugUtilisation")
library(DrugUtilisation)
```

## What is a drug utilisation study?

A drug utilisation study (DUS) describes *how* a drug is used in clinical practice: who receives it, at what dose, for how long, for what indications, and in what combinations.

**DUS are not causal analyses.** They [do not estimate the effect of a drug on an outcome]{.underline}. Their purpose is to describe, or characterise, real-world prescribing patterns. They are a key regulatory tool, regularly commissioned as post-authorisation safety studies (PASS) to monitor whether a drug is being used as intended.

Key questions a DUS answers:

- Who is initiating the drug (demographics, comorbidities)?
- What are the treatment patterns (dose, duration, adherence, switching)?
- What are the indications for use?
- Is the drug being used off-label?
- What are co-prescriptions and concomitant medications?

------------------------------------------------------------------------

## Setup

In this example, we better work with the Eunomia dataset as an example instead of the omock data, because the number and variety of drug exposures available in the synthetic data are more broad in the Eunomia dataset. You can also download example datasets from the `omock` package, like the `GiBleed` dataset. The documentation of the package is [here](https://ohdsi.github.io/omock/index.html).

```{r setup-packages, message=FALSE, warning=FALSE}
library(DrugUtilisation)
library(CohortConstructor)
library(CodelistGenerator)
library(CDMConnector)
library(omopgenerics)
library(omock)
library(dplyr)
library(duckdb)

requireEunomia()
db <- DBI::dbConnect(duckdb::duckdb(), dbdir = eunomiaDir())
cdm <- cdmFromCon(db, 
                  cdmSchema = "main", 
                  writeSchema = "main", 
                  writePrefix = "greg_")


semaglutide <- getDrugIngredientCodes(cdm = cdm,
                                      name = "semaglutide")

t2dm <- getCandidateCodes(
  cdm = cdm,
  domains = "Condition",
  includeDescendants = TRUE, 
  includeAncestor = FALSE,
  keywords = c("type 2 diabetis mellitus", "T2DM")
)

acute_pancreatitis <- getCandidateCodes(
  cdm = cdm,
  domains = "Condition", 
  includeDescendants = FALSE,
  includeAncestor = FALSE,
  keywords = c("Acute pancreatitis")
)

# Build drug exposure cohorts (one per GLP-1 agent)
cdm$semaglutide_cohort <- conceptCohort(
  cdm         = cdm,
  conceptSet  = semaglutide,
  name        = "semaglutide_cohort",   # name of the cohort table in the CDM
  exit = "event_end_date",
  overlap = "merge")


# Build condition cohorts
cdm$t2dm_cohort <- conceptCohort(
  cdm         = cdm,
  conceptSet  = t2dm,
  name        = "t2dm_cohort"
)

cdm$ap_cohort <- conceptCohort(
  cdm         = cdm,
  conceptSet  = acute_pancreatitis,
  name        = "ap_cohort"
)

cdm <- bind(cdm$semaglutide_cohort, cdm$t2dm_cohort, cdm$ap_cohort, name = "study_cohorts")
```

::: callout-note
If the CDM you use does not have the `concept id` that allow you to do the exercises, please, try another CDM or use another substance/disease!
:::

------------------------------------------------------------------------

## Example pipeline

In this tutorial we will provide an example on how to do a DUS with `DrugUtilisation`, but further documentation can be found in the [package documentation](https://darwin-eu.github.io/DrugUtilisation/), with all available functions in its [reference index](https://darwin-eu.github.io/DrugUtilisation/reference/index.html) and in [this other tutorial](https://darwin-eu.github.io/DrugUtilisation/tutorial/#/title-slide) or [this training](https://oxford-pharmacoepi.github.io/HDS-DrugTraining/).

### Step 1: Create a drug cohort

The `DrugUtilisation` package provides `generateIngredientCohortSet()` as its primary cohort creation function. It creates drug cohorts, collapsing consecutive prescription records with gaps below a specified threshold.

```{r create-cohort, message=FALSE, warning=FALSE, eval=FALSE}
# Create semaglutide cohort as ingredient
cdm <- generateIngredientCohortSet(
  cdm        = cdm,
  name       = "semaglutide_dus_cohort",
  ingredient = "semaglutide",
  gapEra     = 365    # this number depends on how recurrent prescriptions for this ingredient's medications are
) 

cdm$semaglutide_dus_cohort <- cdm$semaglutide_dus_cohort|>
  requireIsFirstDrugEntry() |> # First entry, like before
  requireObservationBeforeDrug(days = 365) # Prior observation period, like before!

```

> **Our study:** we have already done our cohorts for GLP-1 RA with the `conceptCohort()` function. There is no need to repeat that. These are two different options: the `generateIngredientCohortSet()` allows you to go for ingredient level, while the `conceptCohort()` allows you to use specific products and dose forms.

```{r q1, echo=FALSE, results='asis'}
cat(
  "**Quick check:** What does `gapEra = 365` mean in `generateIngredientCohortSet()`?\n\n",
  mcq(c(
    "Patients must have 365 days of observation before the drug start",
    answer = "Drug exposure records with fewer than 365 days between the end of one record and the start of the next are collapsed into a single continuous era",
    "The drug cohort is restricted to the first 365 days of treatment",
    "Only prescriptions lasting at least 365 days are included"
  )), "\n\n"
)
```

`r hide("Explanation")` `gapEra` controls the **era collapsing** logic. In claims or dispensing data, a patient on a once-weekly drug might have their 4-week prescription record end on day 28 and the next refill start on day 30, a 2-day gap. Without era collapsing, these would appear as two separate drug eras (treatment discontinuation + restart). With `gapEra = 365`, gaps of fewer than 365 days are bridged, producing one continuous treatment era. The appropriate gap depends on the drug's dosing frequency and the data source's dispensing patterns. `r unhide()`

### Step 2: Treatment duration and exposure

```{r drug-utilisation, message=FALSE, warning=FALSE, eval=FALSE}
dus_result <- cdm$semaglutide_dus_cohort |>
  summariseDrugUtilisation(
    ingredientConceptId = 793143,  # semaglutide
    gapEra              = 365,     # change at convenience
    estimates           = c("q25", "median", "q75", "mean", "sd")
  )

dus_result |>
  filter(variable_name %in% c(
    "days prescribed",
    "days exposed",
    "number eras",
    "initial daily dose milligram"
  )) |>
  select(variable_name, estimate_name, estimate_value)
```

> **Exercise!**
>
> For each GLP-1 RA agent, report:
>
> | Metric | Clinical interpretation |
> |------------------------------------|------------------------------------|
> | Days prescribed (median, IQR) | The duration of the prescribed course |
> | Days exposed (median, IQR) | The duration of actual drug supply based on dispensing records |
> | Number of drug eras | Whether patients have multiple treatment episodes (treatment restarts) |
> | Initial daily dose (mg) | Whether patients are initiated at the recommended starting dose |
> | Cumulative dose (mg) | Total drug exposure over the follow-up period |

### Step 3: Indications for use

What conditions were patients treated for? This is especially relevant for GLP-1 RAs, which are licensed for T2DM and obesity but increasingly used off-label.

```{r indication, message=FALSE, warning=FALSE, eval = F}
# Define indication cohorts
indications <- list(headache  = 378253, influenza = 4266367, t2dm = 201826)

cdm$indications_cohort <- conceptCohort(cdm = cdm,
                                        conceptSet = indications,
                                        name = "indications_cohort")

indication_result <- cdm$dus_cohort |>
  summariseIndication(
    indicationCohortName = "indications_cohort",
    unknownIndicationTable  = "condition_occurrence",
    indicationWindow        = list(c(-30, 0)) # You can select the most appropriate window for your analysis
  )

indication_result |>
  filter(estimate_name == "percentage") |>
  select(variable_level, estimate_value) 
```

```{r tf, echo=FALSE, results='asis'}
cat("**Quick check:** You run the summarise indication function for semaglutide and most of the indications are `unknown`. What is the most plausible cause?\n\n")
cat("- Semaglutide is being used off-label more than what we thought:", torf(FALSE), "\n\n")
cat("- Indication is not well-recorded in the database:", torf(TRUE), "\n\n")
cat("- We have not created a good enough indications cohort:", torf(FALSE), "\n\n")
```

### Step 4: Treatment adherence

The `DrugUtilisation` package allows you to compute treatment adherence in two different ways: - Proportion of Days Covered (PDC): the standard measure of medication adherence in claims data. PDC ≥ 80% is the conventional threshold for "adherent." - As a survival analysis using the `CohortSurvival` package. The function assumes that each cohort entry is a continuous treatment era.

```{r adherence, eval=FALSE}
# Summarise treatment adherence
pdc_adherence_result <- cdm$semaglutide_dus_cohort |>
  summariseProportionOfPatientsCovered(followUpDays = 365)

library(CohortSurvival)
surv_adherence_result <- cdm$semaglutide_dus_cohort |>
  summariseDiscontinuationAsSurvival()

# You can plot the results!
```

### Step 5: Treatment switching and restart

Sometimes, one patient might switch from one treatment to another due to a lot of different causes: secondary effects, treatment failure, economic constraints, new agent introduced in the market... Independently of the reason, treatment switching and restart are valuable information to recover from the data. For our GLP-1 RA DUS, switching is particularly clinically relevant, since patients may switch between agents due to tolerability (nausea is common with all GLP-1 RAs), formulary changes, or lack of glycaemic response. A high rate of early switching (within 90 days) to a different GLP-1 RA would be a DUS finding worth reporting to regulators, as it might indicate a population-level tolerability pattern not captured in clinical trials.

```{r switching, eval=FALSE}
# Summarise switching after treatment discontinuation 
switch_result <- cdm$semaglutide_dus_cohort |>
  summariseDrugRestart(
    switchCohortTable = "GLP1-RA",
    # A cohort that you have had defined with all other GLP1-RA
    followUpDays      = 365
  )

```

------------------------------------------------------------------------

## Exporting the results

You can export a summarised result with all the objects that you created from your different calculations, and then read that file in a Shiny App to show the results. The package for that is [OmopViewer](https://ohdsi.github.io/OmopViewer/) but [visOmopResults](https://darwin-eu.github.io/visOmopResults/) and each package own functions allow for it too.

```{r export, eval=FALSE}
# Suppress small cell counts before sharing
minCellCount <- 5 

exportSummarisedResult(
  dus_result,
  indication_result,
  pdc_adherence_result,
  surv_adherence_result,
  switch_result,
  minCellCount = minCellCount,
  path = here("Results"),
  fileName = "GregonBoard_dus_{cdm_name}_{date}.csv"
)

```
