---
title: "Drug Utilisation"
format: html
---

```{r setup, include=FALSE}
library(webexercises)
install.packages("DrugUtilisation")
library(DrugUtilisation)
```

## What is a drug utilisation study?

A drug utilisation study (DUS) describes *how* a drug is used in clinical practice: who receives it, at what dose, for how long, for what indications, and in what combinations.

**DUS are not causal analyses.** They [do not estimate the effect of a drug on an outcome]{.underline}. Their purpose is to describe, or characterise, real-world prescribing patterns. They are a key regulatory tool, regularly commissioned as post-authorisation safety studies (PASS) to monitor whether a drug is being used as intended.

Key questions a DUS answers:

- Who is initiating the drug (demographics, comorbidities)?
- What are the treatment patterns (dose, duration, adherence, switching)?
- What are the indications for use?
- Is the drug being used off-label?
- What are co-prescriptions and concomitant medications?

------------------------------------------------------------------------

## Setup

```{r setup-packages, message=FALSE, warning=FALSE}
library(DrugUtilisation)
# ------------------------------------------------------------------------------------
# Remember that we previously created semaglutide_cohort, t2dm_cohort and ap_cohort!
# Go to the Cohort Constructor section if you need more info.
```

------------------------------------------------------------------------

## Example pipeline

In this tutorial we will provide an example on how to do a DUS with `DrugUtilisation`, but further documentation can be found in the [package documentation](https://darwin-eu.github.io/DrugUtilisation/), with all available functions in its [reference index](https://darwin-eu.github.io/DrugUtilisation/reference/index.html) and in [this other tutorial](https://darwin-eu.github.io/DrugUtilisation/tutorial/#/title-slide) or [this training](https://oxford-pharmacoepi.github.io/HDS-DrugTraining/).

### Step 1: Create a drug cohort

The `DrugUtilisation` package provides `generateIngredientCohortSet()` as its primary cohort creation function. It creates drug cohorts, collapsing consecutive prescription records with gaps below a specified threshold.

```{r create-cohort, message=FALSE, warning=FALSE}
# Create semaglutide cohort as ingredient
cdm <- generateIngredientCohortSet(
  cdm        = cdm,
  name       = "semaglutide_dus_cohort",
  ingredient = "semaglutide",
  gapEra     = 365    # this number depends on how recurrent prescriptions for this ingredient's medications are
) |>
  requireIsFirstDrugEntry() |> # First entry, like before
  requireObservationBeforeDrug(days = 365) # Prior observation period, like before!

```

> **Our study:** we have already done our cohorts for GLP-1 RA with the `conceptCohort()` function. There is no need to repeat that. These are two different options: the `generateIngredientCohortSet()` allows you to go for ingredient level, while the `conceptCohort()` allows you to use specific products and dose forms.

```{r q1, echo=FALSE, results='asis'}
cat(
  "**Quick check:** What does `gapEra = 365` mean in `generateIngredientCohortSet()`?\n\n",
  mcq(c(
    "Patients must have 365 days of observation before the drug start",
    answer = "Drug exposure records with fewer than 365 days between the end of one record and the start of the next are collapsed into a single continuous era",
    "The drug cohort is restricted to the first 365 days of treatment",
    "Only prescriptions lasting at least 365 days are included"
  )), "\n\n"
)
```

`r hide("Explanation")` `gapEra` controls the **era collapsing** logic. In claims or dispensing data, a patient on a once-weekly drug might have their 4-week prescription record end on day 28 and the next refill start on day 30 — a 2-day gap. Without era collapsing, these would appear as two separate drug eras (treatment discontinuation + restart). With `gapEra = 30`, gaps of fewer than 30 days are bridged, producing one continuous treatment era. The appropriate gap depends on the drug's dosing frequency and the data source's dispensing patterns. `r unhide()`

### Step 2: Treatment duration and exposure

```{r drug-utilisation, message=FALSE, warning=FALSE}
du_result <- cdm$dus_cohort |>
  summariseDrugUtilisation(
    ingredientConceptId = 1125315,  # acetaminophen
    gapEra              = 30,
    estimates           = c("q25", "median", "q75", "mean", "sd")
  )

du_result |>
  filter(variable_name %in% c(
    "days prescribed",
    "days exposed",
    "number eras",
    "initial daily dose milligram"
  )) |>
  select(variable_name, estimate_name, estimate_value)
```

> **Our study — treatment duration metrics to report:**
>
> For each GLP-1 RA agent, report:
>
> | Metric | Clinical interpretation |
> |------------------------------------|------------------------------------|
> | Days prescribed (median, IQR) | The duration of the prescribed course — often longer than actual use |
> | Days exposed (median, IQR) | The duration of actual drug supply based on dispensing records |
> | Number of drug eras | Whether patients have multiple treatment episodes (treatment restarts) |
> | Initial daily dose (mg) | Whether patients are initiated at the recommended starting dose |
> | Cumulative dose (mg) | Total drug exposure over the follow-up period |

### Step 3: Indications for use

What conditions were patients treated for? This is especially relevant for GLP-1 RAs, which are licensed for T2DM and obesity but increasingly used off-label.

```{r indication, message=FALSE, warning=FALSE}
# Define indication cohorts
indications <- list(
  headache  = 378253,
  influenza = 4266367
)

cdm <- CDMConnector::generateConceptCohortSet(
  cdm        = cdm,
  conceptSet = indications,
  name       = "indications_cohort"
)

indication_result <- cdm$dus_cohort |>
  summariseIndication(
    indicationCohortName    = "indications_cohort",
    unknownIndicationTable  = "condition_occurrence",
    indicationWindow        = list(c(-30, 0))
  )

indication_result |>
  filter(estimate_name == "percentage") |>
  select(variable_level, estimate_value)
```

> **Our study — indication analysis:** For GLP-1 RAs, define indication cohorts for:
>
> - **Type 2 diabetes mellitus** (primary licensed indication)
> - **Obesity / overweight** (licensed for semaglutide/tirzepatide in higher doses)
> - **Cardiovascular disease** (licensed for some agents with demonstrated CV benefit)
> - **No identifiable coded indication** (potential off-label use)
>
> A high proportion of "no identifiable coded indication" may reflect coding limitations rather than true off-label use — many GPs do not re-code T2DM at every prescribing event.

### Step 4: Treatment adherence

Proportion of Days Covered (PDC) is the standard measure of medication adherence in claims data. PDC ≥ 80% is the conventional threshold for "adherent."

```{r adherence, eval=FALSE}
# Summarise treatment adherence
adherence_result <- cdm$dus_cohort |>
  summariseTreatmentAdherence(
    followUpDays = 365
  )

adherence_result |>
  filter(variable_name == "Proportion of patients covered") |>
  select(estimate_name, estimate_value)
```

### Step 5: Treatment switching and restart

```{r switching, eval=FALSE}
# Define comparator drug cohorts
cdm <- generateIngredientCohortSet(
  cdm        = cdm,
  name       = "comparator_drugs",
  ingredient = c("ibuprofen", "naproxen")
)

# Summarise switching after treatment discontinuation
switch_result <- cdm$dus_cohort |>
  summariseDrugRestart(
    switchCohortTable = "comparator_drugs",
    followUpDays      = 365
  )

switch_result |>
  filter(variable_name %in% c("restart", "switch", "discontinue")) |>
  select(variable_name, estimate_name, estimate_value)
```

> **Our study — treatment switching:** For GLP-1 RA DUS, switching is particularly clinically relevant: patients may switch between agents due to tolerability (nausea is common with all GLP-1 RAs), formulary changes, or lack of glycaemic response. A high rate of early switching (within 90 days) to a different GLP-1 RA would be a DUS finding worth reporting to regulators, as it might indicate a population-level tolerability pattern not captured in clinical trials.


```{r q2, echo=FALSE, results='asis'}
cat(
  "**Quick check:** Your DUS shows that 35% of semaglutide users have no coded",
  "T2DM or obesity diagnosis in the 30 days before initiation.",
  "Does this mean 35% are receiving the drug off-label?\n\n",
  torf(FALSE), "\n\n"
)
```

`r hide("Explanation")` Not necessarily. In routine clinical databases, many patients with chronic conditions like T2DM do not have their diagnosis re-coded at every GP visit or prescription event. The absence of a recent coded diagnosis in the 30-day window does not mean the diagnosis is absent — it may simply not have been recorded recently. Looking at a wider window (e.g. any T2DM code ever in the patient's history) typically reveals a much higher proportion with documented indication. The "no identifiable coded indication" category in a DUS is therefore best interpreted as a lower bound on true off-label use, not as the actual off-label rate. `r unhide()`

------------------------------------------------------------------------

## Exporting the results

You can export a summarised result with all the objects that you created from your different calculations, and then read that file in a Shiny App to show the results. The package for that is [OmopViewer](https://ohdsi.github.io/OmopViewer/) but [visOmopResults](https://darwin-eu.github.io/visOmopResults/) and each package own functions allow for it too.

```{r export, eval=FALSE}
# Suppress small cell counts before sharing
minCellCount <- 5 

exportSummarisedResult(
  inc,
  prev,
  pointPrev,
  minCellCount = minCellCount,
  path = here("Results"),
  fileName = "GregonBoard_incprev_{cdm_name}_{date}.csv"
)

```

