---
title: "Drug Utilisation"
format: html
---

```{r setup, include=FALSE}
library(webexercises)
install.packages("DrugUtilisation")
library(DrugUtilisation)
```

## What is a drug utilisation study?

A drug utilisation study (DUS) describes *how* a drug is used in clinical practice — who receives it, at what dose, for how long, for what indications, and in what combinations. The World Health Organization defines drug utilisation as "the marketing, distribution, prescription and use of drugs in a society, with special emphasis on the resulting medical, social and economic consequences."

DUS are not causal analyses. They do not estimate the effect of a drug on an outcome. Their purpose is to characterise real-world prescribing patterns, and they are a key regulatory tool: the EMA and national competent authorities regularly commission DUS as post-authorisation safety studies (PASS) to monitor whether a drug is being used as intended.

::: callout-note
## When is a DUS required?

Under EU Regulation 536/2014 and Directive 2001/83/EC, a marketing authorisation holder may be required to conduct a DUS as part of a Risk Management Plan (RMP) or as a post-authorisation obligation. DARWIN EU® has delivered multiple DUS across its network for EMA-mandated studies, all using the OMOP CDM and the `DrugUtilisation` package.
:::

Key questions a DUS answers:

- Who is initiating the drug (demographics, comorbidities)?
- What are the treatment patterns (dose, duration, adherence, switching)?
- What are the indications for use?
- Is the drug being used off-label?
- What are co-prescriptions and concomitant medications?

------------------------------------------------------------------------

## Setup

```{r setup-packages, message=FALSE, warning=FALSE}
library(DrugUtilisation)
library(CohortCharacteristics)
library(CDMConnector)
library(omopgenerics)
library(dplyr)

# Synthetic CDM
cdm <- mockDrugUtilisation(numberIndividual = 500)
```

------------------------------------------------------------------------

## Step 1: Create a drug cohort

The `DrugUtilisation` package provides `generateIngredientCohortSet()` as its primary cohort creation function. It creates drug era cohorts, collapsing consecutive prescription records with gaps below a specified threshold.

```{r create-cohort, message=FALSE, warning=FALSE}
# Create a cohort of acetaminophen users
# (proxy for our GLP-1 RA cohort in the real study)
cdm <- generateIngredientCohortSet(
  cdm        = cdm,
  name       = "dus_cohort",
  ingredient = "acetaminophen",
  gapEra     = 30    # collapse records with <30-day gaps into one era
)

# Restrict to first episode (new-user design)
cdm$dus_cohort <- cdm$dus_cohort |>
  requireIsFirstDrugEntry() |>
  requireObservationBeforeDrug(days = 365)

cohortCount(cdm$dus_cohort)
```

> **Our study — creating GLP-1 RA drug cohorts:** In the real analysis, we use a gap era of 30 days for once-weekly agents (semaglutide, dulaglutide, tirzepatide) and 7 days for twice-daily agents (exenatide immediate release). This reflects the expected prescription patterns and avoids artificially splitting continuous treatment episodes due to minor dispensing gaps.

```{r q1, echo=FALSE, results='asis'}
cat(
  "**Quick check:** What does `gapEra = 30` mean in `generateIngredientCohortSet()`?\n\n",
  mcq(c(
    "Patients must have 30 days of observation before the drug start",
    answer = "Drug exposure records with fewer than 30 days between the end of one record and the start of the next are collapsed into a single continuous era",
    "The drug cohort is restricted to the first 30 days of treatment",
    "Only prescriptions lasting at least 30 days are included"
  )), "\n\n"
)
```

`r hide("Explanation")` `gapEra` controls the **era collapsing** logic. In claims or dispensing data, a patient on a once-weekly drug might have their 4-week prescription record end on day 28 and the next refill start on day 30 — a 2-day gap. Without era collapsing, these would appear as two separate drug eras (treatment discontinuation + restart). With `gapEra = 30`, gaps of fewer than 30 days are bridged, producing one continuous treatment era. The appropriate gap depends on the drug's dosing frequency and the data source's dispensing patterns. `r unhide()`

------------------------------------------------------------------------

## Step 2: Treatment duration and exposure

```{r drug-utilisation, message=FALSE, warning=FALSE}
du_result <- cdm$dus_cohort |>
  summariseDrugUtilisation(
    ingredientConceptId = 1125315,  # acetaminophen
    gapEra              = 30,
    estimates           = c("q25", "median", "q75", "mean", "sd")
  )

du_result |>
  filter(variable_name %in% c(
    "days prescribed",
    "days exposed",
    "number eras",
    "initial daily dose milligram"
  )) |>
  select(variable_name, estimate_name, estimate_value)
```

> **Our study — treatment duration metrics to report:**
>
> For each GLP-1 RA agent, report:
>
> | Metric | Clinical interpretation |
> |------------------------------------|------------------------------------|
> | Days prescribed (median, IQR) | The duration of the prescribed course — often longer than actual use |
> | Days exposed (median, IQR) | The duration of actual drug supply based on dispensing records |
> | Number of drug eras | Whether patients have multiple treatment episodes (treatment restarts) |
> | Initial daily dose (mg) | Whether patients are initiated at the recommended starting dose |
> | Cumulative dose (mg) | Total drug exposure over the follow-up period |

------------------------------------------------------------------------

## Step 3: Indications for use

What conditions were patients treated for? This is especially relevant for GLP-1 RAs, which are licensed for T2DM and obesity but increasingly used off-label.

```{r indication, message=FALSE, warning=FALSE}
# Define indication cohorts
indications <- list(
  headache  = 378253,
  influenza = 4266367
)

cdm <- CDMConnector::generateConceptCohortSet(
  cdm        = cdm,
  conceptSet = indications,
  name       = "indications_cohort"
)

indication_result <- cdm$dus_cohort |>
  summariseIndication(
    indicationCohortName    = "indications_cohort",
    unknownIndicationTable  = "condition_occurrence",
    indicationWindow        = list(c(-30, 0))
  )

indication_result |>
  filter(estimate_name == "percentage") |>
  select(variable_level, estimate_value)
```

> **Our study — indication analysis:** For GLP-1 RAs, define indication cohorts for:
>
> - **Type 2 diabetes mellitus** (primary licensed indication)
> - **Obesity / overweight** (licensed for semaglutide/tirzepatide in higher doses)
> - **Cardiovascular disease** (licensed for some agents with demonstrated CV benefit)
> - **No identifiable coded indication** (potential off-label use)
>
> A high proportion of "no identifiable coded indication" may reflect coding limitations rather than true off-label use — many GPs do not re-code T2DM at every prescribing event.

------------------------------------------------------------------------

## Step 4: Treatment adherence

Proportion of Days Covered (PDC) is the standard measure of medication adherence in claims data. PDC ≥ 80% is the conventional threshold for "adherent."

```{r adherence, eval=FALSE}
# Summarise treatment adherence
adherence_result <- cdm$dus_cohort |>
  summariseTreatmentAdherence(
    followUpDays = 365
  )

adherence_result |>
  filter(variable_name == "Proportion of patients covered") |>
  select(estimate_name, estimate_value)
```

------------------------------------------------------------------------

## Step 5: Treatment switching and restart

```{r switching, eval=FALSE}
# Define comparator drug cohorts
cdm <- generateIngredientCohortSet(
  cdm        = cdm,
  name       = "comparator_drugs",
  ingredient = c("ibuprofen", "naproxen")
)

# Summarise switching after treatment discontinuation
switch_result <- cdm$dus_cohort |>
  summariseDrugRestart(
    switchCohortTable = "comparator_drugs",
    followUpDays      = 365
  )

switch_result |>
  filter(variable_name %in% c("restart", "switch", "discontinue")) |>
  select(variable_name, estimate_name, estimate_value)
```

> **Our study — treatment switching:** For GLP-1 RA DUS, switching is particularly clinically relevant: patients may switch between agents due to tolerability (nausea is common with all GLP-1 RAs), formulary changes, or lack of glycaemic response. A high rate of early switching (within 90 days) to a different GLP-1 RA would be a DUS finding worth reporting to regulators, as it might indicate a population-level tolerability pattern not captured in clinical trials.

------------------------------------------------------------------------

## Combine and share results

```{r combine, message=FALSE, warning=FALSE}
# all_results <- omopgenerics::bind(
#   du_result,
#   indication_result
# ) |>
#   omopgenerics::suppress(minCellCount = 5)
# 
# write.csv(all_results, "results/drug_utilisation.csv", row.names = FALSE)
```

```{r q2, echo=FALSE, results='asis'}
cat(
  "**Quick check:** Your DUS shows that 35% of semaglutide users have no coded",
  "T2DM or obesity diagnosis in the 30 days before initiation.",
  "Does this mean 35% are receiving the drug off-label?\n\n",
  torf(FALSE), "\n\n"
)
```

`r hide("Explanation")` Not necessarily. In routine clinical databases, many patients with chronic conditions like T2DM do not have their diagnosis re-coded at every GP visit or prescription event. The absence of a recent coded diagnosis in the 30-day window does not mean the diagnosis is absent — it may simply not have been recorded recently. Looking at a wider window (e.g. any T2DM code ever in the patient's history) typically reveals a much higher proportion with documented indication. The "no identifiable coded indication" category in a DUS is therefore best interpreted as a lower bound on true off-label use, not as the actual off-label rate. `r unhide()`
