---
title: "Estimating effectiveness"
format: html
---

```{r setup, include=FALSE}
library(webexercises)
```

## From description to causal inference

All the analyses in previous chapters were descriptive or exploratory. They are used to characterise populations, estimate rates, and describe drug utilisation. An effectiveness (or comparative safety) study attempts to **estimate a causal effect** of one treatment compared to another.

::: callout-note
The standard is to perform "comparative" studies (one treatment vs. another) instead of comparing to no treatment. This is to more closely reproduce a clinical trial framework and avoid selection bias.
:::

Assessing causality is the hardest task in pharmacoepidemiology. Confounding is the central threat. In a randomised trial, random allocation ensures that confounders are distributed equally between groups. In observational data, we must adjust for confounding analytically, using tools like propensity scores, but we are restricted to the variables available in the data.

The preferred approach to causal inference from observational data is the **new-user active-comparator cohort design with propensity score adjustment and negative control calibration**.

::: callout-tip
In your [protocol](study_protocol.qmd), you used the ICH E9(R1) estimand framework to define your research question and objectives. Keep that framework in mind during this chapter!
:::

------------------------------------------------------------------------

## Setup

In this example, we better work with the Eunomia dataset as an example instead of the omock data, because the number and variety of drug exposures available in the synthetic data are more broad in the Eunomia dataset. You can also download example datasets from the `omock` package, like the `GiBleed` dataset. The documentation of the package is [here](https://ohdsi.github.io/omock/index.html).

::: callout-caution
In this setup section, we present a more realistic construction of cohorts introducing exitAtDeath() and relevant intersections with condition cohorts.
:::

```{r setup-packages, message=FALSE, warning=FALSE, eval=FALSE}
library(CohortSurvival)
library(CohortMethod)
library(CohortConstructor)
library(CodelistGenerator)
library(CDMConnector)
library(omopgenerics)
library(omock)
library(dplyr)
library(duckdb)

requireEunomia()
db <- DBI::dbConnect(duckdb::duckdb(), dbdir = eunomiaDir())
cdm <- cdmFromCon(db, 
                  cdmSchema = "main", 
                  writeSchema = "main", 
                  writePrefix = "greg_")


semaglutide <- getDrugIngredientCodes(cdm = cdm,
                                      name = "semaglutide")

tirzepatide <- getDrugIngredientCodes(cdm = cdm,
                                      name = "tirzepatide")

t2dm <- getCandidateCodes(
  cdm = cdm,
  domains = "Condition",
  includeDescendants = TRUE, 
  includeAncestor = FALSE,
  keywords = c("type 2 diabetis mellitus", "T2DM")
)

acute_pancreatitis <- getCandidateCodes(
  cdm = cdm,
  domains = "Condition", 
  includeDescendants = FALSE,
  includeAncestor = FALSE,
  keywords = c("Acute pancreatitis")
)

# Build exposure cohorts (one per GLP-1 agent)
cdm$semaglutide_cohort <- conceptCohort(
  cdm = cdm,
  conceptSet = semaglutide,
  name = "semaglutide_cohort",
  exit = "event_end_date",
  overlap = "merge"
)

cdm$tirzepatide_cohort <- conceptCohort(
  cdm = cdm,
  conceptSet = tirzepatide,
  name = "tirzepatide_cohort",
  exit = "event_end_date",
  overlap = "merge"
)

# Build the T2DM cohort
cdm$t2dm_cohort <- conceptCohort(
  cdm = cdm,
  conceptSet = t2dm,
  name = "t2dm_cohort"
)

# Keep only the first qualifying exposure record for each person.
cdm$semaglutide_cohort <- requireIsFirstEntry(
  cdm$semaglutide_cohort,
  name = "semaglutide_cohort"
)

cdm$tirzepatide_cohort <- requireIsFirstEntry(
  cdm$tirzepatide_cohort,
  name = "tirzepatide_cohort"
)

# Keep only GLP-1 users who also have T2DM.
# Each exposure is intersected with T2DM separately.
cdm$semaglutide_t2dm <- requireCohortIntersect(
  cohort = cdm$semaglutide_cohort,
  targetCohortTable = "t2dm_cohort",
  window = list(c(-Inf, 0)),
  intersections = c(1, Inf),
  name = "semaglutide_t2dm"
)

cdm$tirzepatide_t2dm <- requireCohortIntersect(
  cohort = cdm$tirzepatide_cohort,
  targetCohortTable = "t2dm_cohort",
  window = list(c(-Inf, 0)),
  intersections = c(1, Inf),
  name = "tirzepatide_t2dm"
)

# Build the acute pancreatitis outcome cohort
cdm$ap_cohort <- conceptCohort(
  cdm = cdm,
  conceptSet = acute_pancreatitis,
  name = "ap_cohort"
)

# Exit at death; otherwise keep the existing cohort end date.
cdm$semaglutide_t2dm <- exitAtDeath(
  cdm$semaglutide_t2dm,
  name = "semaglutide_t2dm"
)

cdm$tirzepatide_t2dm <- exitAtDeath(
  cdm$tirzepatide_t2dm,
  name = "tirzepatide_t2dm"
)

# Combine the two exposure cohorts into one study target table.
cdm <- bind(
  cdm$semaglutide_t2dm,
  cdm$tirzepatide_t2dm,
  name = "study_cohorts"
)


## We need a special connection for CohortMethod -------------
library(DatabaseConnector)
library(FeatureExtraction)
install.packages("Eunomia")
library(Eunomia)

data_dir <- file.path(tempdir(), "eunomia")
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

Eunomia::downloadEunomiaData(
  datasetName = "GiBleed",
  pathToData = data_dir,
  overwrite = TRUE
)

list.files(data_dir, recursive = TRUE)

db_file <- tempfile(fileext = ".sqlite")

Eunomia::extractLoadData(
  from = file.path(data_dir, "GiBleed_5.3.zip"),
  to = db_file,
  dbms = "sqlite",
  verbose = TRUE
)


# Connection information
connectionDetails <- DatabaseConnector::createConnectionDetails(dbms = "sqlite", server = db_file)

# OMOP CDM schema
cdmDatabaseSchema <- "cdm"
cohortDatabaseSchema <- "my_results"
cohortTable <- "my_cohorts"

library(Capr)
semaglutideConceptId <- 793143
tirzepatideConceptId <- 779705
t2dmConceptId <- 201826
acutePancreatitisConceptId <- 199074

semaglutide_cl <- cs(descendants(semaglutideConceptId), name = "semaglutide")
semaglutideCohort <- cohort(entry = entry(drugExposure(semaglutide_cl)),
                            exit = exit(
                              endStrategy = drugExit(
                                semaglutide_cl,
                                persistenceWindow = 365,
                                surveillanceWindow = 0
                              )
                            ))

tirzepatide_cl <- cs(descendants(tirzepatideConceptId), name = "tirzepatide")
tirzepatideCohort <- cohort(entry = entry(drugExposure(tirzepatide_cl)),
                            exit = exit(
                              endStrategy = drugExit(
                                tirzepatide_cl,
                                persistenceWindow = 365,
                                surveillanceWindow = 0
                              )
                            ))

t2dm_cl <- cs(descendants(t2dmConceptId), name = "Type 2 Diabetes")
t2dmCohort <- cohort(
  entry = entry(conditionOccurrence(t2dm_cl, firstOccurrence())),
  exit = exit(endStrategy = observationExit())
)

acutePancreatitis_cl <- cs(descendants(acutePancreatitisConceptId), name = "Acute Pancreatitis")
acutePancreatitisCohort <- cohort(
  entry = entry(conditionOccurrence(acutePancreatitis_cl, firstOccurrence())),
  exit = exit(endStrategy = observationExit())
)

exposuresAndIndicationCohorts <- makeCohortSet(semaglutideCohort,
                                               tirzepatideCohort,
                                               t2dmCohort,
                                               acutePancreatitisCohort)

library(CohortGenerator)
cohortTableNames <- getCohortTableNames(cohortTable = cohortTable)
createCohortTables(
  connectionDetails = connectionDetails,
  cohortDatabaseSchema = cohortDatabaseSchema,
  cohortTableNames = cohortTableNames
)
generateCohortSet(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = cdmDatabaseSchema,
  cohortDatabaseSchema = cohortDatabaseSchema,
  cohortTableNames = cohortTableNames,
  cohortDefinitionSet = exposuresAndIndicationCohorts
)
```

------------------------------------------------------------------------

## Example pipeline

This chapter uses a single comparative effectiveness question throughout:

> **Among people with T2DM who initiate semaglutide or tirzepatide, what is the comparative risk of acute pancreatitis?**

The workflow is intentionally short:

1.  Build semaglutide and tirzepatide exposure cohorts.
2.  Intersect each exposure with T2DM.
3.  Use the first qualifying exposure as the index date.
4.  Exit follow-up at acute pancreatitis, death, observation end, or 3 years.
5.  Extract baseline covariates.
6.  Estimate propensity scores.
7.  Match on the propensity score.
8.  Estimate the hazard ratio with Cox regression.
9.  Repeat with IPTW as a sensitivity analysis.
10. Plot Kaplan-Meier curves.
11. Use negative controls to assess systematic error.
12. Extract the final results.

`CohortSurvival` is used for the descriptive time-to-event/Kaplan-Meier part, while `CohortMethod` is used for propensity scores, matching, IPTW, and Cox regression.

### Step 1: Define the study population

Our question is:

> **Among people with T2DM who newly start semaglutide or tirzepatide, does the risk of acute pancreatitis differ between treatments?**

We use a **new-user, active-comparator cohort design**:

- **Target:** semaglutide + T2DM
- **Comparator:** tirzepatide + T2DM
- **Outcome:** acute pancreatitis
- **Index date:** first qualifying exposure date
- **Prior outcome:** exclude people with previous acute pancreatitis
- **Follow-up:** from treatment initiation until acute pancreatitis, death, the end of observation, or the end of the study risk window

The intersection with T2DM was created in the setup section. `exitAtDeath()` makes death an explicit censoring point for the exposure cohorts. `CohortConstructor` supports both cohort intersections and changing cohort end dates at death.

#### CohortSurvival

`CohortSurvival` can estimate Kaplan-Meier survival directly from the target and outcome cohorts. The target cohort defines who is at risk and the outcome cohort defines the event.

```{r step1-survival, eval=FALSE}
survivalResult <- estimateSingleEventSurvival(
  cdm = cdm,
  targetCohortTable = "study_cohorts",
  outcomeCohortTable = "ap_cohort",
  outcomeWashout = Inf,       # no previous AP
  followUpDays = 1095,        # 3 years
  censorOnCohortExit = TRUE,  # includes death exit
  strata = list("cohort_definition_id")
)
```

#### CohortMethod

For causal estimation, `CohortMethod` creates the study population from the two exposure cohorts and the outcome. It can exclude people with prior outcomes and define the risk window.

```{r step1-method, eval=FALSE}
# Extract the data needed for CohortMethod -----------------------------
# Define which types of covariates must be constructed:
covSettings <- createDefaultCovariateSettings(
  excludedCovariateConceptIds = c(semaglutideConceptId, tirzepatideConceptId),
  addDescendantsToExclude = TRUE
)

# Load data:
cohortMethodData <- getDbCohortMethodData(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = cdmDatabaseSchema,
  targetId = 1,
  comparatorId = 2,
  outcomeIds = 4,
  exposureDatabaseSchema = cohortDatabaseSchema,
  exposureTable = cohortTable,
  outcomeDatabaseSchema = cohortDatabaseSchema,
  outcomeTable = cohortTable,
  nestingCohortDatabaseSchema = cohortDatabaseSchema,
  nestingCohortTable = cohortTable,
  getDbCohortMethodDataArgs = createGetDbCohortMethodDataArgs(
    removeDuplicateSubjects = "keep first, truncate to second",
    firstExposureOnly = TRUE,
    washoutPeriod = 365,
    restrictToCommonPeriod = TRUE,
    nestingCohortId = 3,
    covariateSettings = covSettings
  )
)
cohortMethodData

studyPop <- createStudyPopulation(
  cohortMethodData = cohortMethodData,
  removeSubjectsWithPriorOutcome = TRUE,
  minDaysAtRisk = 1,
  riskWindowStart = 0,
  startAnchor = "cohort start",
  riskWindowEnd = 1095,
  endAnchor = "cohort end"
)

```

For this study, baseline variables should include information such as:

- age and sex
- T2DM history and duration
- obesity and related metabolic conditions
- previous pancreatitis or pancreatic disease
- gallbladder disease
- alcohol-related conditions
- hypertriglyceridaemia
- renal disease
- cardiovascular disease
- diabetes medications and other relevant drugs
- healthcare utilisation

::: callout-important
Simple rule for covariates: **use information measured before treatment starts**.
:::

### Step 2: Estimate the propensity score

The propensity score summarises the probability of receiving treatment A (semaglutide) rather than treatment B (tirzepatide), given measured baseline covariates. `CohortMethod::createPs()` fits the propensity model. In current versions of `CohortMethod`, IPTW is also derived from the propensity scores.

#### CohortSurvival

`CohortSurvival` uses a weight column when estimating the Kaplan-Meier curve to account for the Propensity Score, so you obtained the weighted survival directly.

```{r step3-survival-weight, eval=FALSE}
weightedSurvival <- estimateSingleEventSurvival(
  cdm = cdm,
  targetCohortTable = "study_cohorts",
  outcomeCohortTable = "ap_cohort",
  weight = "iptw",
  outcomeWashout = Inf,
  followUpDays = 1095,
  censorOnCohortExit = TRUE
)
```

#### CohortMethod

```{r step3-ps, eval=FALSE}
ps <- createPs(
  cohortMethodData = cohortMethodData,
  population = studyPop
)

computePsAuc(ps)

plotPs(ps,
targetLabel = "semaglutide",
comparatorLabel = "tirzepatide",
showCountsLabel = TRUE,
showAucLabel = TRUE,
showEquipoiseLabel = TRUE)
```

::: callout-tip
Look for **substantial overlap** in the propensity-score distributions. Poor overlap means that some patients are very unlikely to receive one of the two treatments, so the comparison becomes dependent on extrapolation.
:::

```{r q1, echo=FALSE, results='asis'}
cat(
  "**Quick check:** If the two propensity-score distributions have almost no overlap, what problem does this create for a causal comparison?\n\n",
  mcq(c(
    "You can do the Cox regression without adjustment for confounding because there is no confounding if the populations are too different.",
    answer = "The causal effect estimate will be unreliable because the treatment groups are not comparable.",
    "The causal effect estimate is unreliable because the propensity scores were calculated incorrectly."
  )), "\n\n"
)
```

### Step 3: Adjust for confounding

Here we show two simple approaches:

1.  **Propensity-score matching**
2.  **Inverse probability of treatment weighting (IPTW)**

#### Propensity-score matching

```{r step4-matching, eval=FALSE}
matchedPop <- matchOnPs(
  population = ps,
  matchOnPsArgs = createMatchOnPsArgs(
    maxRatio = 1,
    caliper = 0.2, # adjust at convenience
    caliperScale = "standardized logit"
  )
)

balance <- computeCovariateBalance(
  population = matchedPop,
  cohortMethodData = cmData
)

plotCovariateBalanceScatterPlot(
  balance,
  showMaxLabel = TRUE
)
```

::: callout-note
After matching, baseline covariates should be reasonably balanced between semaglutide and tirzepatide.
:::

`matchOnPs()` performs propensity-score matching and removes people who cannot be matched within the specified criteria.

#### IPTW

```{r step4-iptw, eval=FALSE}
iptwPop <- truncateIptw(
  population = ps,
  truncateIptwArgs = createTruncateIptwArgs(
    maxWeight = 10
  )
)
```

Truncation reduces the influence of very large weights. `CohortMethod` supports IPTW directly in the outcome model.

```{r tf, echo=FALSE, results='asis'}
cat("Why might matching and IPTW give slightly different estimates even when both use the same propensity score?")
cat("They must produce exactly the same estimate because they use the same propensity score.", torf(FALSE))
cat("Matching focuses on creating comparable treated-control pairs, whereas IPTW uses the propensity score to weight the entire sample, so the resulting target populations can differ.", torf(TRUE))
cat("IPTW can give very large weights to patients with propensity scores near 0 or 1, while matching may exclude patients who lack suitable counterparts.", torf(TRUE))
cat("Any difference between matching and IPTW proves that the propensity-score model was incorrectly specified.", torf(FALSE))
```

### Step 4: Estimate the effect of treatment with Cox regression

The main outcome is time to first acute pancreatitis.

#### CohortMethod: Cox regression after matching

```{r step5-cox, eval=FALSE}
outcomeModel <- fitOutcomeModel(
  population = matchedPop,
  fitOutcomeModelArgs = createFitOutcomeModelArgs(
    modelType = "cox",
    stratified = TRUE
  )
)

```

::: callout-note
The main result is the **hazard ratio (HR)** for semaglutide versus tirzepatide.

- HR \< 1: lower instantaneous risk with semaglutide
- HR = 1: no difference
- HR \> 1: higher instantaneous risk with semaglutide
:::

`CohortMethod` uses Cox regression for time-to-event analyses and can fit the model after propensity-score matching.

#### IPTW Cox regression

```{r step5-iptw-cox, eval=FALSE}
iptwOutcomeModel <- fitOutcomeModel(
  population = iptwPop,
  fitOutcomeModelArgs = createFitOutcomeModelArgs(
    modelType = "cox",
    inversePtWeighting = TRUE,
    bootstrapCi = TRUE
  )
)

```

::: callout-important
A hazard ratio is a relative measure. It does not tell us the absolute probability of acute pancreatitis. The Kaplan-Meier analysis in the next step helps describe the absolute event experience over time.
:::

```{r tf2, echo=FALSE, results='asis'}
cat("Suppose the HR for semaglutide versus tirzepatide is 0.80 with a 95% CI of 0.60-1.07. Is it correct to say that semaglutide reduces acute pancreatitis risk?")
cat("Yes because the average effect is under 1.", torf(FALSE))
cat("No, because the confidence interval crosses 1, the threshold for signifcant difference of effect.", torf(TRUE))
cat("No, but because you need to add that reduces the comparative risk of pancreatitis.", torf(FALSE))
```

`r hide("Explanation")` Although you do need to word it as "comparative risk" because you are not obtaining a measure of the absolute risk of acute pancreatitis but the comparative risk of suffering from acute pancreatitis depending on the treatment you take! `r unhide()`

### Step 5: Kaplan-Meier curves

Kaplan-Meier curves provide a simple visual description of time to acute pancreatitis.

#### CohortSurvival

```{r step6-km-survival, eval=FALSE}
survivalResult <- estimateSingleEventSurvival(
  cdm = cdm,
  targetCohortTable = "study_cohorts",
  outcomeCohortTable = "ap_cohort",
  outcomeWashout = Inf,
  followUpDays = 1095,
  censorOnCohortExit = TRUE,
  strata = list("cohort_definition_id")
)

# or, as you saw before!

weightedSurvival <- estimateSingleEventSurvival(
  cdm = cdm,
  targetCohortTable = "study_cohorts",
  outcomeCohortTable = "ap_cohort",
  weight = "iptw",
  outcomeWashout = Inf,
  followUpDays = 1095,
  censorOnCohortExit = TRUE,
  strata = list("cohort_definition_id")
)

```

The result contains survival estimates and event information that can be used to produce Kaplan-Meier displays.

::: callout-note
`CohortSurvival` also supports weighting and stratification.
:::

#### CohortMethod

```{r step6-km-method, eval=FALSE}
plotKaplanMeier(
  population = matchedPop,
  outcomeId = acutePancreatitisId,
  confidenceIntervals = TRUE
)
```

The two curves should be interpreted together with the Cox estimate. A small visual separation does not necessarily imply a clinically important difference.

### Negative controls

A negative control is an outcome that you include in the analysis for which there is **no plausible causal effect of the treatment**, but which is recorded in the same healthcare data. Including negative control outcomes is good praxis to look for evidence of systematic error such as residual confounding.

::: callout-warning
Do not choose negative controls just because they are easy to find. They should be clinically unrelated to the exposure and outcome, while still being recorded often enough to be informative.
:::

### Extracting the results

The final extraction should contain only the results needed to report this study:

- number of semaglutide users
- number of tirzepatide users
- number with acute pancreatitis
- follow-up time
- matched Cox HR and 95% CI
- IPTW Cox HR and 95% CI
- Kaplan-Meier estimates
- negative-control estimates

For the survival results, `CohortSurvival` returns a `summarised_result` object that can be inspected or converted with `asSurvivalResult()`.

```{r extract-results, eval=FALSE}
# Survival results
survivalTable <- asSurvivalResult(
  survivalResult
)

# Main Cox estimate
mainEstimate <- data.frame(
  analysis = "PS matched Cox",
  estimate = coef(outcomeModel),
  lower_95 = confint(outcomeModel)[1],
  upper_95 = confint(outcomeModel)[2]
)

# IPTW sensitivity estimate
iptwEstimate <- data.frame(
  analysis = "IPTW Cox",
  estimate = coef(iptwOutcomeModel),
  lower_95 = confint(iptwOutcomeModel)[1],
  upper_95 = confint(iptwOutcomeModel)[2]
)

mainEstimate
iptwEstimate
survivalTable
```

::: callout-tip
For a study report, keep the extraction step separate from the analysis code. This makes it easier to rerun the analysis without changing how results are reported.
:::

## Exporting the results

You can export a summarised result with all the objects that you created from your different calculations, and then read that file in a Shiny App to show the results. The package for that is [OmopViewer](https://ohdsi.github.io/OmopViewer/) but [visOmopResults](https://darwin-eu.github.io/visOmopResults/) and each package own functions allow for it too.

```{r export, eval=FALSE}
# Suppress small cell counts before sharing
minCellCount <- 5 

exportSummarisedResult(
  mainEstimate,
  iptwEstimate,
  survivalTable,
  minCellCount = minCellCount,
  path = here("Results"),
  fileName = "GregonBoard_survival_{cdm_name}_{date}.csv"
)

```
