---
title: "CDM using R"
format: html
---

```{r setup, include=FALSE}
library(webexercises)
install.packages("CDMConnector")
install.packages("dplyr")
install.packages("omock")
library(CDMConnector)
library(dplyr)
library(omock)
```

## Why R for OMOP?

ATLAS is a powerful graphical interface, but it has limits: complex cohort manipulations, custom statistical models, and some automated multi-study pipelines all might require additional `code`. The OHDSI community and the DARWIN EU Coordination Centre have co-developed a rich ecosystem of R packages — collectively known as the **OMOPverse** — that allow you to work directly with OMOP CDM data in R in a consistent, reproducible, and database-agnostic way.

The key architectural principle of the OMOPverse is that **you never need to change your analysis code when you change your database**. Whether your data lives in DuckDB, PostgreSQL, SQL Server, Snowflake, or Redshift, the same R functions work identically because the packages use `dbplyr` to translate R code into database-specific SQL at runtime.

::: callout-tip
You can check what are SQL databases [here](https://cloud.google.com/discover/what-are-sql-databases). And find a global OMOP CDM analytics tutorial [here](https://dpa-pde-oxford.quarto.pub/omop-cdm-analytics/#/title-slide).
:::

::: callout-note
## The OMOPverse core packages

The packages covered across these tutorial chapters are:

| Package | Purpose |
|------------------------------------|------------------------------------|
| [`CDMConnector`](https://darwin-eu.github.io/CDMConnector/index.html) | Connect to any OMOP CDM database |
| [`omopgenerics`](https://darwin-eu.github.io/omopgenerics/) | Shared data structures and utilities |
| [`omock`](https://ohdsi.github.io/omock/) | Generate synthetic OMOP data for testing |
| [`CohortDiagnostics`](https://ohdsi.github.io/CohortDiagnostics/index.html) | Generate a set of diagnostic features for cohorts |
| [`PhenotypeR`](https://ohdsi.github.io/PhenotypeR/index.html) | Generate a set of diagnostic features for cohorts |
| [`CodelistGenerator`](https://darwin-eu.github.io/CodelistGenerator/) | Create concept sets to build cohorts in R |
| [`CohortConstructor`](https://ohdsi.github.io/CohortConstructor/) | Build and manipulate cohorts in R |
| [`CohortCharacteristics`](https://darwin-eu.github.io/CohortCharacteristics/) | Summarise cohort characteristics |
| [`IncidencePrevalence`](https://darwin-eu.github.io/IncidencePrevalence/) | Estimate incidence rates and prevalence |
| [`DrugUtilisation`](https://darwin-eu.github.io/DrugUtilisation/) | Drug cohort creation and DUS analyses |
| [`CohortSurvival`](https://darwin-eu-dev.github.io/CohortSurvival/) | Causal comparative effectiveness (DARWIN EU®) |
| [`CohortMethod`](https://ohdsi.github.io/CohortMethod/) | Causal comparative effectiveness (OHDSI) |

All are available either on CRAN or downloadable from their GitHub repository.
:::

------------------------------------------------------------------------

## Installation

### Installing R and RStudio

Please, check this guide: <https://ohdsi.github.io/Hades/rSetup.html>

### Setting up R

Ideally, when you are doing R code for a GREG study, you will create a new R project and set up the environment. This includes installing all the packages required for the project and recording the package versions in a `.lock` file. By doing so, you can restore the exact same environment whenever you open the project in a new session or on a different computer, ensuring that the code remains reproducible and consistent over time.

A common workflow is to use the **renv** package, which creates an isolated project library and manages package dependencies. After installing the required packages, you can save the environment with `renv::snapshot()`, which generates the `renv.lock` file. Later, anyone working on the project can recreate the same environment by running `renv::restore()`. This approach minimizes compatibility issues caused by package updates and helps guarantee that analyses can be reproduced exactly as originally performed.

Install all packages you will need for these tutorials and set up the environment:

```{r install, eval=F}
install.packages("renv")

renv::init()

install.packages(c(
  "CDMConnector",
  "omopgenerics",
  "omock",
  "CohortDiagnostics",
  "PhenotypeR",
  "CodelistGenerator",
  "CohortConstructor",
  "CohortCharacteristics",
  "IncidencePrevalence",
  "DrugUtilisation",
  "CohortSurvival",
  "CohortMethod",
  "dplyr",
  "duckdb",
  "DBI",
  "ggplot2"
))

renv::snapshot()
```

Later, when you (or someone else) open the project on a different computer or in a fresh R session, simply run:

```{r, eval=FALSE}
renv::restore()
```

::: callout-tip
A good practice is to include the following commands at the beginning of your project setup script:

```{r, eval=FALSE}
# ------------------------------------------------------------
# Project setup
# ------------------------------------------------------------

# Install renv if it is not available
if (!requireNamespace("renv", quietly = TRUE)) {
  install.packages("renv")
}

# Restore the project environment
renv::activate()
renv::restore()
```

**But!** Notice that some databases might **not** be connected to the internet and therefore you will not be able to recover the environment with these commands. In that case, you will need to manually install all packages required for the study!

```{r install-local, eval=F}
# If the package is in a local folder:
install.packages("C:/path to folder with the package", 
                 repos = NULL, 
                 type = "source")
```
:::

::: callout-caution
Always remember to load all installed packages necessary for your project with the "library" command!

```{r load, eval = F}
library(CDMConnector)
library(omopgenerics)
library(omock)
library(CohortDiagnostics)
library(PhenotypeR)
library(CodelistGenerator)
# [...]
```
:::

### Working on GitHub

To make the project easier to share, collaborate on, and keep track of changes, it is recommended to use GitHub as a version-control platform. The first step is to create a new repository on GitHub. A repository acts as the central location where all project files, scripts, documentation, and version history are stored.

To create a new repository, log in to your GitHub account, select **New repository**, choose a descriptive name for the project, and initialize it with a README file if desired. Once the repository has been created, clone it to your local computer using either Git or the GitHub interface. The local copy of the repository will contain all project files and will be connected to the remote GitHub repository.

::: callout-tip
You can use [GitHub Desktop](https://desktop.github.com/download/), it is very convenient!
:::

After cloning the repository, open the project directly from the local folder in RStudio by selecting **File → New Project → Existing Directory** and choosing the repository folder. This creates an R project (`.Rproj`) inside the GitHub repository, allowing RStudio to recognize the project structure and keep all files organized.

Once the project is set up, changes can be tracked using Git. After modifying files or adding new scripts, commit the changes with a meaningful message and push them to GitHub. This workflow ensures that the project history is preserved and that collaborators can access the most recent version of the analysis.

For reproducible research, the GitHub repository should include the R project file, the analysis scripts, documentation files, and the `renv.lock` file created with `renv::snapshot()`. The `renv.lock` file allows collaborators to recreate the exact R environment required to run the project.

------------------------------------------------------------------------

## Connecting to an OMOP CDM

With all set up, let's enter the OMOPverse! To all OMOP analyses, the entrance door is the **CDM reference**, an R object that represents your database connection and knows where your OMOP tables live (see [OMOP CDM](OMOPCDM.qmd)). This object in R is created with `CDMConnector`. You can either work with a real database or synthetic data. If you have access to a real database, you can connect to that one. Nevertheless, some analyses might be very computationally expensive and spend days running.

::: callout-tip
You might want to use synthetic data to try your code before running it on real data!
:::

### A. Real database (e.g. PostgreSQL)

```{r real-db, eval=FALSE}
library(CDMConnector)
library(DBI)
library(RPostgres)

# 1. Create a database connection

# 1.1 Database metadata and connection details -----

# name of the database, use acronym in capital letters 
dbName <-

# 1.2 parameters to connect to create cdm object ----

# name of the schema where cdm tables are located
cdmSchema <- "cdm"
 
# name of a schema in the database where you have writing permission
writeSchema <- 

# name of the achilles schema
achillesSchema <- 

# combination of at least 5 letters + _ (eg. "abcde_") that will lead any table
# written in the write schema
writePrefix <- 

# connection
con <- DBI::dbConnect(RPostgres::Postgres(), # Go to: https://dbi.r-dbi.org/
                      host = db_host,
                      port = db_port,
                      dbname = db_cdmvers,
                      user = db_user,
                      password = db_password)

cdm <- cdmFromCon(con = con,
                  cdmSchema = cdmSchema,
                  writeSchema = writeSchema,
                  achillesSchema = achillesSchema,
                  cdmName = "your_study_name")

cdm
```

::: callout-warning
Use `usethis::edit_r_environ()` to open your `.Renviron` file and store connection credentials there. This prevents accidentally committing passwords to Git and makes your code portable across team members.
:::

### B. Synthetic data with `omock` (what we use in this tutorial)

For all code examples in these chapters, we use synthetic data generated by the `omock` package. This means you can run every code block yourself, right now, without needing a real database connection. You can also do this when planning a study, to avoid running unnecessary long codes or codes with bugs in real databases.

```{r mock-cdm, eval=F}
library(omopgenerics)
library(omock)
library(CDMConnector)
library(dplyr)

# Create a synthetic CDM with 1000 people
cdm <- omock::mockCdmReference() |>
  omock::mockPerson(nPerson = 1000) |>
  omock::mockObservationPeriod() |>
  omock::mockConditionOccurrence() |>
  omock::mockDrugExposure() |>
  omock::mockDeath()

cdm
```

## Exploring the CDM structure

Once you have a CDM reference, you can inspect it like a named list. Each element is a [lazy database table](https://search.r-project.org/CRAN/refmans/dtplyr/html/lazy_dt.html), it does not load data into memory until you explicitly `collect()` it.

```{r explore-cdm, eval=FALSE}
# List all available tables
names(cdm)

# Inspect the person table
cdm$person |> glimpse()

# Count rows without loading everything into memory
cdm$person |> count()

# Look at a few drug exposure records
cdm$drug_exposure |> head(10) |> 
  collect() # At this moment is when we actually retrieve the data!
```

::: callout-note
When you write `cdm$drug_exposure |> filter(drug_concept_id == 1503297)`, R does **not** immediately execute this against the database. Instead, it builds a query that is only sent to the database when you call `collect()`, `count()`, `glimpse()`, or a `summarise*` function. This means you can build complex multi-step queries without pulling millions of rows into memory.
:::

```{r lazy-demo, eval=FALSE}
# This builds a query but does NOT run it yet
query <- cdm$condition_occurrence |>
  filter(condition_concept_id == 201826) |>  # Type 2 diabetes
  select(person_id, condition_start_date)

# This sends the query to the database and returns results
query |> collect() |> head(5)
```

## The `omopgenerics` result format

All OMOPverse analytical functions return results in a standardised `summarised_result` format: a tibble with a consistent set of columns regardless of which analysis was run. Understanding this format is essential because all downstream table and plot functions expect it.

```{r result-format, eval=FALSE}
# A summarised_result always has these columns:
# result_id, cdm_name, group_name, group_level,
# strata_name, strata_level, variable_name, variable_level,
# estimate_name, estimate_type, estimate_value,
# additional_name, additional_level

# You can check if an object is a summarised_result
omopgenerics::isSummarisedResult(my_result)

# Suppress small cell counts before sharing (required for privacy)
my_result |> omopgenerics::suppress(minCellCount = 5)
```

::: callout-important
The `suppress(minCellCount = 5)` function replaces any count below 5 with `<5` and sets related percentages to `NA`. This is a minimum standard for privacy protection in federated network studies. **You should always apply it before exporting results from a data partner.**
:::

## Disconnecting

Always close your database connection at the end of a session:

```{r disconnect, eval=FALSE}
# For a real database connection
CDMConnector::cdmDisconnect(cdm)
# or 
dbDisconnect(con)

# For a mock CDM created with omock
omopgenerics::cdmDisconnect(cdm)
# or
dbDisconnect(con)
```

## More information

You can expand your knowledge on CDM Connector through this [tutorial](https://dpa-pde-oxford.quarto.pub/omop-and-cdmconnector), or [this](https://ohdsi.github.io/Tidy-R-programming-with-OMOP/omop/cdm_reference.html) online book with coding exercises. The latter, the [Tidy R programming with OMOP](https://ohdsi.github.io/Tidy-R-programming-with-OMOP/omop/cdm_reference.html) book, will be a good guide for the following chapters as well. As always, you can additionally check [The Book of OHDSI](https://ohdsi.github.io/TheBookOfOhdsi/SqlAndR.html).
